---
title: "Supporting 10 common mistakes article"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/10mistakes

## Introduction

```{r,libs}

#knitr::opts_chunk$set(dev = 'svg') # set output device to svg

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("eulerr")
  library("gplots")
  library("mitch")
  library("fgsea")
  library("RhpcBLASctl")
  library("parallel")
  library("beeswarm")
  library("vioplot")
})

RhpcBLASctl::blas_set_num_threads(1)

if ( ! dir.exists("img") ) {
  dir.create("img")
}

```

For this guide I will be using bulk RNA-seq data from a previous study, which is
deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.

The raw data have been processed by the DEE2 project, and the summary gene expression counts are
available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
(Ziemann et al, 2019).

The gene counts have also been deposited to the `/data` folder in the `example.Rdata` file in case
the DEE2 resource becomes unavailable.
To import it, use the command: `load("data/example.Rdata")`

Alternatively, you could fetch data from another resource like NCBI GEO, Zenodo or from the host
storage drive.

```{r,fetch}

mdat <- getDEE2Metadata("hsapiens")

# get sample sheet
ss <- subset(mdat,SRP_accession=="SRP038101")

# fetch the whole set of RNA-seq data
x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
mx <- x$GeneCounts
rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
dim(mx)

# aza no filtering
ss$trt <- grepl("Treated",ss$Experiment_title)

ss %>%
  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
  kable_paper("hover", full_width = F)

```

## Data quality control

QC is important, even if you are using public transcriptome data.
For RNA-seq it is a good idea to show the number of reads for each sample.

```{r,counts1}

par(mar=c(5,7,5,1))
barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")

```

Now make a MDS plot.

```{r,mds1}

mds <- cmdscale(dist(t(mx)))

# expand plot area
XMIN=min(mds[,1])*1.3
XMAX=max(mds[,1])*1.3
YMIN=min(mds[,2])*1.3
YMAX=max(mds[,2])*1.3

cols <- as.character(grepl("Treated",ss$Experiment_title))
cols <- gsub("FALSE","lightblue",cols)
cols <- gsub("TRUE","pink",cols)
plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
  pch=19,cex=2,col=cols, main="MDS plot")
text(cmdscale(dist(t(mx))), labels=colnames(mx) )

```

## Differential expression analysis

I will be using DESeq2 for DE analysis, the count matrix is prefiltered using
a detection threshold of 10 reads per sample across all samples.
All genes that meet the detection threshold will comprise the background list.
The first 6 rows of the count matrix are shows.

```{r,de1}

mxf <- mx[which(rowMeans(mx)>=10),]
dim(mxf)

head(mxf,6) %>%
  kbl(caption="Count matrix format") %>%
  kable_paper("hover", full_width = F)

dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <-cbind(as.data.frame(z),mxf)
def <-as.data.frame(zz[order(zz$pvalue),])

head(def,10) %>%
  kbl(caption="Top DE genes for Aza treatment") %>%
  kable_paper("hover", full_width = F)

```

Make a smear plot.

```{r,smearplot}

sigf <- subset(def,padj<=0.05)

DET=nrow(mxf)
NSIG=nrow(sigf)
NUP=nrow(subset(sigf,log2FoldChange>0))
NDN=nrow(subset(sigf,log2FoldChange<0))

HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")

plot(log10(def$baseMean) ,def$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="5-azacitidine treatment in AML3 cells",
  xlab="log10(basemean)",ylab="log2(fold change)")

points(log10(sigf$baseMean) ,sigf$log2FoldChange,
  cex=0.6,pch=19,col="red")

mtext(HEADER)

```

In the next sections I will run enrichment analysis with over-representation test
and compare it to functional class scoring.

## Get update gene symbols

Let's see if many of the gene symbols have changed.

Comparing Ensembl release 90 to GENCODE Release 49 (GRCh38.p14) which corresponds to Ensembl 115 (Sept 2025).

So v90 had 58k genes and v115 has 86k which is a big increase.
The v90 has 1609 genes that have been removed an v115 has 29676 that have been added.
There are 56693 common.

From these common ones, 20456 have different gene names.
I think the approach to naming these putative genes has changed a lot.
Lots of CXorfXX genes have changed, not just AL358472.5 type names.

In the next chunk, I'll show a few lines of the older source gene names.

There is also an Euler diagram of the gene symbol repertoires.

```{r,genesymbolupdate,fig.height=3,fig.width=5}

gt <- read.table("https://dee2.io/data/hsapiens/hsa_gene_info.tsv",header=TRUE)
head(gt)

gtnew <- read.table("ref/gencode.v49.genenames.tsv")

v1 <- list("Ens90"=gt$GeneID, "Ens115"=gtnew$V1)
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

```

```{r,genesymbolupdate2,fig.height=3,fig.width=5}

message("Number of unique genes in the original gene table:")
length(gt$GeneID)

message("Number of unique genes in the new gene annotation set:")
length(gtnew$V1)

message("Number of shared gene IDs between new and old annotations:")
length(intersect(gt$GeneID,gtnew$V1))

message("Number of gene IDs specific to the old annotation:")
length(setdiff(gt$GeneID,gtnew$V1))

message("Number of gene IDs specific to the new annotation:")
length(setdiff(gtnew$V1,gt$GeneID))

message("Here is the mapping table for old-new gene symbols:")
gtm <- merge(gt,gtnew,by.x="GeneID",by.y="V1")

message("Number of rows:")
nrow(gtm)

message("How many of these gene symbols are the same?")
table(gtm$GeneSymbol == gtm$V2)

```

## Get gene sets

Reading in Reactome gene sets.

```{r,gs}
# from https://reactome.org/download/current/ReactomePathways.gmt.zip
gsets <- gmtPathways("ref/ReactomePathways_2025-09-02.gmt")

```

## Mistake 1 - Raw p-values

First, discard old gene symbols in favour of the new ones.

In this analysis, we are doing ORA with fora, and comparing the results
of raw p-values versus using FDR correction.

This corresponds to the results shown in Figure 1B.

```{r,mistake1,fig.height=3,fig.width=5}

up <- rownames(subset(def, log2FoldChange>0 & padj<0.05))
up <- sapply(strsplit(up," "),'[[',1)
up <- unique(gtnew[match(up,gtnew$V1),2])
up <- up[which(!is.na(up))]

dn <- rownames(subset(def, log2FoldChange<0 & padj<0.05))
dn <- sapply(strsplit(dn," "),'[[',1)
dn <- unique(gtnew[match(dn,gtnew$V1),2])
dn <- dn[which(!is.na(dn))]

bg <- rownames(def)
bg <- sapply(strsplit(bg," "),'[[',1)
bg <- unique(gtnew[match(bg,gtnew$V1),2])
bg <- bg[which(!is.na(bg))]

message("length of up, down and background sets")
lapply(list(up,dn,bg),length)

fres_up <- fora(pathways=gsets, genes=up, universe=bg, minSize = 5, maxSize = Inf)

message("number of gene sets in the results")
nrow(fres_up)

message("number of up-regulated gene sets with FDR<0.05")
nrow(subset(fres_up,padj<0.05))
message("number of up-regulated gene sets with p<0.05")
nrow(subset(fres_up,pval<0.05))

fres_up_fdr <- subset(fres_up,padj<0.05)$pathway
fres_up_nom <- subset(fres_up,pval<0.05)$pathway

fres_dn <- fora(pathways=gsets, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
message("number of down-regulated gene sets with FDR<0.05")
nrow(subset(fres_dn,padj<0.05))
message("number of down-regulated gene sets with p<0.05")
nrow(subset(fres_dn,pval<0.05))

fres_dn_fdr <- subset(fres_dn,padj<0.05)$pathway
fres_dn_nom <- subset(fres_dn,pval<0.05)$pathway

v1 <- list("up p"=fres_up_nom, "up FDR"=fres_up_fdr,
  "dn p"=fres_dn_nom, "dn FDR"=fres_dn_fdr)

par(mar=c(1,1,1,1))
par(mfrow=c(1,1))
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

fres_fdr <- union(fres_dn_fdr,fres_up_fdr)
fres_nom <- union(fres_dn_nom,fres_up_nom)

v1 <- list("p"=fres_nom, "FDR"=fres_fdr)
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

unlink("img/1b.png")
png("img/1b.png", width = 480, height = 400)
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
dev.off()
par(mfrow=c(1,1))

unlink("img/1b.pdf")
pdf("img/1b.pdf", width = 5, height = 5)
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
dev.off()
par(mfrow=c(1,1))

par(mar=c(5.1,4.1,4.1,2.1))

```

That analysis shows how a big difference between FDR and raw p-values,
but it doesn't prove that the p-value hits are false positives.

Here, we can prove they are false positives by randomly selecting 1000
genes from the list of detected ones, and testing them for enrichment.

When doing enrichment analysis of a random list of 1000 genes, we got an
average of 51.6 pathways with p<0.05, while after p-value correction,
that number was just 0.16 with FDR<0.05.
This correponds to the data shown in Figure 1A.

```{r,mistake1_sim,fig.height=6,fig.width=6}

seeds <- seq(from=100,to=100*100,by=100)

m1sim <- mclapply(seeds, function(seed) {
  set.seed(seed)
  rand <- sample(x=bg,size=1000,replace=FALSE)
  fres_rand <- fora(pathways=gsets, genes=rand, universe=bg, minSize = 5, maxSize = Inf)
  nom <- nrow(subset(fres_rand,pval<0.05))
  fdr <- nrow(subset(fres_rand,padj<0.05))
  c(nom,fdr)
},mc.cores=8)

m1sim <- do.call(rbind,m1sim)
colnames(m1sim) <- c("p","FDR")

boxplot(m1sim,main="Random gene sets (100 repetitions)",ylab="no. gene sets meeting \"significance\" threshold <0.05",frame=FALSE)
beeswarm(list(mean(m1sim[,1]), mean(m1sim[,2])),add=TRUE,pch=4,cex=2)
lab1 <- paste("Mean=",signif(mean(m1sim[,1]) , 3),sep="")
lab2 <- paste("Mean=",signif(mean(m1sim[,2]) , 2),sep="")
text(c(1.2,2.2), c( mean(m1sim[,1])+5, mean(m1sim[,2])+5 ), labels=c(lab1,lab2) )

unlink("img/1a.png")
png("img/1a.png", width = 480, height = 480)
boxplot(m1sim,main="Random gene sets (100 repetitions)",ylab="no. gene sets meeting \"significance\" threshold <0.05",frame=FALSE)
beeswarm(list(mean(m1sim[,1]), mean(m1sim[,2])),add=TRUE,pch=4,cex=2)
lab1 <- paste("Mean=",signif(mean(m1sim[,1]) , 3),sep="")
lab2 <- paste("Mean=",signif(mean(m1sim[,2]) , 2),sep="")
text(c(1.2,2.2), c( mean(m1sim[,1])+5, mean(m1sim[,2])+5 ), labels=c(lab1,lab2) )
dev.off()
par(mfrow=c(1,1))

unlink("img/1a.pdf")
pdf("img/1a.pdf", width = 5, height = 5)
boxplot(m1sim,main="Random gene sets (100 repetitions)",ylab="no. gene sets meeting \"significance\" threshold <0.05",frame=FALSE)
beeswarm(list(mean(m1sim[,1]), mean(m1sim[,2])),add=TRUE,pch=4,cex=2)
lab1 <- paste("Mean=",signif(mean(m1sim[,1]) , 3),sep="")
lab2 <- paste("Mean=",signif(mean(m1sim[,2]) , 2),sep="")
text(c(1.2,2.2), c( mean(m1sim[,1])+5, mean(m1sim[,2])+5 ), labels=c(lab1,lab2) )
dev.off()
par(mfrow=c(1,1))

par(mar=c(5.1,4.1,4.1,2.1))

```

## Mistake 2 - Background

Will try FORA.

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | FORA (check fgsea version at foot of report) |
| Statistical test used | hypergeometric |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection | DESeq2 FDR<0.05 and log2FC>0 or log2FC<0 |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 5 |

```{r,detectionthreshold}

message("Number of genes in the count matrix")
nrow(x$GeneCounts)

rowmeans <- rowMeans(x$GeneCounts)

message("Number of genes in the count matrix with more than 10 reads per gene on average.")
length(rowmeans[which(rowmeans>=10)])

message("Number of genes in the count matrix with less than 10 reads per gene on average.")
length(rowmeans[which(rowmeans<10)])

message("Proportion of genes below the detection threshold.")
length(rowmeans[which(rowmeans<10)]) / length(rowmeans)

message("Proportion of genes above the detection threshold.")
( 1- sum(rowmeans[which(rowmeans>=10)]) / sum(rowmeans) ) * 100

unfilt <- x$GeneCounts

message("Run deseq2 for differential expression with and without filtering.")
dds_unfilt <- DESeqDataSetFromMatrix(countData = unfilt , colData = ss, design = ~ trt )
res_unfilt <- DESeq(dds_unfilt)
z_unfilt <- results(res_unfilt)
vsd_unfilt <- vst(dds_unfilt, blind=FALSE)
zz_unfilt <-cbind(as.data.frame(z_unfilt),unfilt)
def_unfilt <-as.data.frame(zz_unfilt[order(zz_unfilt$pvalue),])

filt <- unfilt[which(rowMeans(unfilt)>=10),]
dds_filt <- DESeqDataSetFromMatrix(countData = filt , colData = ss, design = ~ trt )
res_filt <- DESeq(dds_filt)
z_filt <- results(res_filt)
vsd_filt <- vst(dds_filt, blind=FALSE)
zz_filt <-cbind(as.data.frame(z_filt),filt)
def_filt <-as.data.frame(zz_filt[order(zz_filt$pvalue),])

message("No significant genes without filtering low genes")
nrow(subset(def_unfilt,padj<0.05))

message("No significant genes with filtering low genes")
nrow(subset(def_filt,padj<0.05))

```

The violin chart is used to understand whether the distribution of gene expression levels
of the background gene list (NS) is similar to the distributrion of DEGs.
If there is a big difference, it is bad, because we already know that genes with high
expression in a cell type are likely to be very different ontologies when compared to lower
expressed genes.
In the unfiltered pair of violin charts, we see that there is a big difference in the
distributions indicating that a lot of the enrichment results could be solely due to
basemean expression levels, not related to the up or down regulation at all.
When filtering is applied, we see that the difference between DEG and NS distributions is
smaller.

```{r,viochart}

deg_unfilt <- subset(def_unfilt,padj<0.05)
ns_unfilt <- subset(def_unfilt,padj>0.05)

rowmeans_deg_unfilt <- rowMeans(deg_unfilt[,7:12])
rowmeans_ns_unfilt <- rowMeans(ns_unfilt[,7:12])

viodata <- list("DEG"=log10(rowmeans_deg_unfilt+0.01),"NS"=log10(rowmeans_ns_unfilt+0.01))

deg_filt <- subset(def_filt,padj<0.05)
ns_filt <- subset(def_filt,padj>0.05)

rowmeans_deg_filt <- rowMeans(deg_filt[,7:12])
rowmeans_ns_filt <- rowMeans(ns_filt[,7:12])

message("Distribution of gene expression levels")
viodata2 <- list("DEG unfilt"=log10(rowmeans_deg_unfilt+0.01),
  "NS unfilt"=log10(rowmeans_ns_unfilt+0.01),
  "DEG filt"=log10(rowmeans_deg_filt+0.01),
  "NS filt"=log10(rowmeans_ns_filt+0.01))

vioplot(viodata2,ylab="log10 mean expression",las=2,side="left")

lapply(viodata2,summary)

```

### Mistake 2 - example dataset analysis

In this next analysis, we run parallel fora tests, one uses the correct
background made up of detected genes (bg), while the other one
uses all genes in the genome (marked with *).

The second euler diagram shown below corresponds to Figure 2C.

```{r,mistake2,fig.height=3,fig.width=5}

# up dn and bg are defined above
all <- unique(gtm$V2)

message("no. genes in each of the lists up, down, bg and all.")
lapply(list(up,dn,bg,all),length)

fres_up_bg <- fora(pathways=gsets, genes=up, universe=bg, minSize = 5, maxSize = Inf)
fres_up_bg <- subset(fres_up_bg,padj<0.05)$pathway
message("no up-regulated pathways with the correct background.")
length(fres_up_bg)

fres_up_all <- fora(pathways=gsets, genes=up, universe=all, minSize = 5, maxSize = Inf)
fres_up_all <- subset(fres_up_all,padj<0.05)$pathway
message("no up-regulated pathways with wg background.")
length(fres_up_all)

fres_dn_bg <- fora(pathways=gsets, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
fres_dn_bg <- subset(fres_dn_bg,padj<0.05)$pathway
message("no down-regulated pathways with the correct background.")
length(fres_dn_bg)

fres_dn_all <- fora(pathways=gsets, genes=dn, universe=all, minSize = 5, maxSize = Inf)
fres_dn_all <- subset(fres_dn_all,padj<0.05)$pathway
message("no down-regulated pathways with wg background.")
length(fres_dn_all)

v1 <- list("up"=fres_up_bg, "up*"=fres_up_all, "dn"=fres_dn_bg, "dn*"=fres_dn_all)
message("no. significant pathways with correct and incorrect* background with direction specified.")
par(mar=c(1,1,1,1))
par(mfrow=c(1,1))
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

v2 <- list("bg"=union(fres_up_bg,fres_dn_bg),
  "bg*"=union(fres_up_all, fres_dn_all) )

message("no. significant pathways with correct and incorrect* background. Direction not specified.")
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2))

par(mar=c(1,1,1,1))
unlink("img/2c.png")
png("img/2c.png", width = 480, height = 400)
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2))
dev.off()
par(mfrow=c(1,1))

par(mar=c(1,1,1,1))
unlink("img/2c.pdf")
pdf("img/2c.pdf", width = 5, height = 5)
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2))
dev.off()
par(mfrow=c(1,1))

par(mar=c(5.1,4.1,4.1,2.1))

```

### Mistake 2 - simulation analysis

Here, we select genes randomly from the genes above the detection threshold (bg list) and
compare to whole genome background.

There are ~444 gene sets significant on average with this approach, as the genes expressed
in AML3 cells are distinctive compared to the whole genome annotation.

When using the correct background, a mean of 0.16 was obtained. This shows that false
positives can be nearly eliminated by using this detection threshold.

This result corresponds to Figure 2A.

```{r,mistake2sim,fig.height=6,fig.width=6}

seeds <- seq(from=100,to=100*100,by=100)

m2sim <- mclapply(seeds, function(seed) {
  set.seed(seed)
  rand <- sample(x=bg,size=1000,replace=FALSE)
  sim_all <- fora(pathways=gsets, genes=rand, universe=all, minSize = 5, maxSize = Inf)
  sim_bg <- fora(pathways=gsets, genes=rand, universe=bg, minSize = 5, maxSize = Inf)
  sig_all <- nrow(subset(sim_all,padj<0.05))
  sig_bg <- nrow(subset(sim_bg,padj<0.05))
  c(sig_all,sig_bg)
},mc.cores=8)

m2sim <- do.call(rbind,m2sim)
colnames(m2sim) <- c("Whole genome background","Correct background")

par(mar=c(5.1,4.1,4.1,2.1))

boxplot(m2sim,main="Random gene sets drawn from BG",ylab="no. gene sets FDR<0.05",frame=FALSE)
mtext("100 simulations")
beeswarm(list(mean(m2sim[,1]), mean(m2sim[,2])),add=TRUE,pch=4,cex=2)
lab1 <- paste("Mean=",signif(mean(m2sim[,1]) , 3),sep="")
lab2 <- paste("Mean=",signif(mean(m2sim[,2]) , 2),sep="")
text(c(1.2,2.2), c( mean(m2sim[,1])+15, mean(m2sim[,2])+15 ), labels=c(lab1,lab2) )

par(mar=c(5.1,4.1,4.1,2.1))
unlink("img/2a.png")
png("img/2a.png", width = 480, height = 480)
boxplot(m2sim,main="Random gene sets drawn from BG",ylab="no. gene sets FDR<0.05",frame=FALSE)
mtext("100 simulations")
beeswarm(list(mean(m2sim[,1]), mean(m2sim[,2])),add=TRUE,pch=4,cex=2)
lab1 <- paste("Mean=",signif(mean(m2sim[,1]) , 3),sep="")
lab2 <- paste("Mean=",signif(mean(m2sim[,2]) , 2),sep="")
text(c(1.2,2.2), c( mean(m2sim[,1])+15, mean(m2sim[,2])+15 ), labels=c(lab1,lab2) )
dev.off()
par(mfrow=c(1,1))

par(mar=c(5.1,4.1,4.1,2.1))
unlink("img/2a.pdf")
pdf("img/2a.pdf", width = 5, height = 5)
boxplot(m2sim,main="Random gene sets drawn from BG",ylab="no. gene sets FDR<0.05",frame=FALSE)
mtext("100 simulations")
beeswarm(list(mean(m2sim[,1]), mean(m2sim[,2])),add=TRUE,pch=4,cex=2)
lab1 <- paste("Mean=",signif(mean(m2sim[,1]) , 3),sep="")
lab2 <- paste("Mean=",signif(mean(m2sim[,2]) , 2),sep="")
text(c(1.2,2.2), c( mean(m2sim[,1])+15, mean(m2sim[,2])+15 ), labels=c(lab1,lab2) )
dev.off()
par(mfrow=c(1,1))

par(mar=c(5.1,4.1,4.1,2.1))

```

In the simulation, which gene sets falsely come up most?

There's 37 pathways that came up 100/100 simulations.
This corresponds to Figure 2B.

```{r,m2sim2,fig.height=6,fig.width=6}

m2simpw <- mclapply(seeds, function(seed) {
  set.seed(seed)
  rand <- sample(x=bg,size=1000,replace=FALSE)
  sim_all <- fora(pathways=gsets, genes=rand, universe=all, minSize = 5, maxSize = Inf)
  sig_all <- subset(sim_all,padj<0.05)$pathway
  return(sig_all)
},mc.cores=8)

m2tbl <- table(unlist((m2simpw)))
m2pw <- table(m2tbl)

plot(table(m2tbl),type="b",
  xlab="no. appearances as false positive, from 100 simulations",
  ylab="no. gene sets",
  main="1000 random genes drawn from expressed genes",
  bty="n"
)
set.seed(42) ; m2selected <- sample(names(which(m2tbl==100)),20)
text(100,seq(from=40,to=120,by=(120-40)/20),labels=m2selected,cex=0.8,pos=2)

par(mar=c(5.1,4.1,4.1,2.1))
unlink("img/2b.png")
png("img/2b.png", width = 480, height = 480)
plot(table(m2tbl),type="b",
  xlab="no. appearances as false positive, from 100 simulations",
  ylab="no. gene sets",
  main="1000 random genes drawn from expressed genes",
  bty="n"
)
set.seed(42) ; m2selected <- sample(names(which(m2tbl==100)),20)
text(100,seq(from=40,to=120,by=(120-40)/20),labels=m2selected,cex=0.8,pos=2)
dev.off()
par(mfrow=c(1,1))

par(mar=c(5.1,4.1,4.1,2.1))
unlink("img/2b.pdf")
pdf("img/2b.pdf", width = 5, height = 5)
plot(table(m2tbl),type="b",
  xlab="no. appearances as false positive, from 100 simulations",
  ylab="no. gene sets",
  main="1000 random genes drawn from expressed genes",
  bty="n"
)
set.seed(42) ; m2selected <- sample(names(which(m2tbl==100)),20)
text(100,seq(from=40,to=120,by=(120-40)/20),labels=m2selected,cex=0.7,pos=2)
dev.off()
par(mfrow=c(1,1))

```

## Mistake 3 and 4 - not reporting enrichment scores and prioritisation only by p-values

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | mitch (check version at foot of report) |
| Statistical test used | Rank ANOVA (Kruskal Wallis) |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Scoring | DESeq2 Wald stat |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 5 |

This scatterplot shows how prioritising by FDR can lead to missing the gene sets with the
biggest effect sizes.
This corresponds to Figure 3.

```{r,mistake4a}

def2 <- def
rownames(def2) <- sapply(strsplit(rownames(def)," "),'[[',1)

m <- mitch_import(x=def2,DEtype="deseq2",geneTable=gtnew)
mres <- mitch_calc(x=m,genesets=gsets,minsetsize=5,cores=4,priority="significance")

message("No. significant gene sets")
nrow(subset(mres$enrichment_result,p.adjustANOVA<0.05))
message("No. gene sets included")
nrow(subset(mres$enrichment_result,p.adjustANOVA<1))

mresdf <- mres$enrichment_result

sig <- subset(mresdf,p.adjustANOVA<0.05)

plot(abs(mresdf$s.dist),-log10(mresdf$pANOVA),
  main="enrichment scores versus significance",
  bty="n",pch=19,cex=1,col="darkgray",
  xlab="absolute enrichment score",
  ylab="-log10(p-value)",
  xlim=c(0,0.95))
points(abs(sig$s.dist),-log10(sig$pANOVA),
  pch=19,cex=1,col="red")
points(c(0.35,0.8),c(42,8),cex=17,pch=0)
text(c(0.35,0.8),c(42,8),c("generic, boring","specific,\ninteresting"))

par(mar=c(5.1,4.1,4.1,2.1))
unlink("img/3a.png")
png("img/3a.png", width = 480, height = 480)
plot(abs(mresdf$s.dist),-log10(mresdf$pANOVA),
  main="enrichment scores versus significance",
  bty="n",pch=19,cex=1,col="darkgray",
  xlab="absolute enrichment score",
  ylab="-log10(p-value)",
  xlim=c(0,0.95))
points(abs(sig$s.dist),-log10(sig$pANOVA),
  pch=19,cex=1,col="red")
points(c(0.35,0.8),c(41,8),cex=17,pch=0)
text(c(0.35,0.8),c(41,8),c("generic, boring","specific,\ninteresting"))
dev.off()
par(mfrow=c(1,1))
par(mar=c(5.1,4.1,4.1,2.1))

par(mar=c(5.1,4.1,4.1,2.1))
unlink("img/3a.pdf")
pdf("img/3a.pdf", width = 5, height = 5)
plot(abs(mresdf$s.dist),-log10(mresdf$pANOVA),
  main="enrichment scores versus significance",
  bty="n",pch=19,cex=1,col="darkgray",
  xlab="absolute enrichment score",
  ylab="-log10(p-value)",
  xlim=c(0,0.95))
points(abs(sig$s.dist),-log10(sig$pANOVA),
  pch=19,cex=1,col="red")
points(c(0.35,0.8),c(41,8),cex=17,pch=0)
text(c(0.35,0.8),c(41,8),c("generic, boring","specific,\ninteresting"))
dev.off()
par(mfrow=c(1,1))
par(mar=c(5.1,4.1,4.1,2.1))

```

Now, let's take a look at the top results when prioritising with FDR only and with
enrichment score.

FDR prioritisation emphasises larger and more generic functions.

ES prioritisation emphasises smaller, more specific categories with larger effect sizes.

To prove this, I go back to the DE table and fetch the log2 fold changes.

These tables correspond to Table 1 and 2.

```{r,mistake4a_table}

mres_sig_up <- head(subset(mres$enrichment_result,s.dist>0),10)
mres_sig_dn <- head(subset(mres$enrichment_result,s.dist<0),10)

# get log2fc values
mres_sig_up$mean_lfc <- sapply(mres_sig_up$set[1:10], function(y) {
  genes <- gsets[[which(names(gsets) == y)]]
  gid <- unique(gtnew[match(genes,gtnew$V2),1])
  lfc <- def2[which(rownames(def2) %in% gid),"log2FoldChange"]
  return(mean(lfc))
})

mres_sig_dn$mean_lfc <- sapply(mres_sig_dn$set[1:10], function(y) {
  genes <- gsets[[which(names(gsets) == y)]]
  gid <- unique(gtnew[match(genes,gtnew$V2),1])
  lfc <- def2[which(rownames(def2) %in% gid),"log2FoldChange"]
  return(mean(lfc))
})

mres_sig_up_tbl <- apply(data.frame(mres_sig_up[,1:2],
  apply(mres_sig_up[,3:6],2,signif,2)),2,as.character)

mres_sig_dn_tbl <- apply(data.frame(mres_sig_dn[,1:2],
  apply(mres_sig_dn[,3:6],2,signif,2)),2,as.character)

mres_sig_up_tbl %>%
  kbl(caption="Top upregulated pathways when prioritised by FDR. ",row.names=FALSE) %>%
  kable_paper("hover", full_width = F)

mres_sig_dn_tbl %>%
  kbl(caption="Top downregulated pathways when prioritised by FDR",row.names=FALSE) %>%
  kable_paper("hover", full_width = F)

mres <- mitch_calc(x=m,genesets=gsets,minsetsize=5,cores=4,priority="effect")

mres_es_up <- head(subset(mres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05),10)
mres_es_dn <- head(subset(mres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05),10)

# get log2fc values
mres_es_up$mean_lfc <- sapply(mres_es_up$set[1:10], function(y) {
  genes <- gsets[[which(names(gsets) == y)]]
  gid <- unique(gtnew[match(genes,gtnew$V2),1])
  lfc <- def2[which(rownames(def2) %in% gid),"log2FoldChange"]
  return(mean(lfc))
})

mres_es_dn$mean_lfc <- sapply(mres_es_dn$set[1:10], function(y) {
  genes <- gsets[[which(names(gsets) == y)]]
  gid <- unique(gtnew[match(genes,gtnew$V2),1])
  lfc <- def2[which(rownames(def2) %in% gid),"log2FoldChange"]
  return(mean(lfc))
})

mres_es_up_tbl <- apply(data.frame(mres_es_up[,1:2],
  apply(mres_es_up[,3:6],2,signif,2)),2,as.character)

mres_es_dn_tbl <- apply(data.frame(mres_es_dn[,1:2],
  apply(mres_es_dn[,3:6],2,signif,2)),2,as.character)

mres_es_up_tbl %>%
  kbl(caption="Top upregulated pathways when prioritised by ES (discarding any with FDR>0.05)",row.names=FALSE) %>%
  kable_paper("hover", full_width = F)

mres_es_dn_tbl %>%
  kbl(caption="Top upregulated pathways when prioritised by ES (discarding any with FDR>0.05)",row.names=FALSE) %>%
  kable_paper("hover", full_width = F)

m4tbls <- list("esup"=mres_es_up_tbl,"esdn"=mres_es_dn_tbl,"sigup"=mres_sig_up_tbl,"sigdn"=mres_sig_dn_tbl)
saveRDS(m4tbls,"m4tbls.Rds")


message("Are mean log2FC values for ES prioritisation bigger than FDR approach?")

message("Are mean log2FC values for ES prioritisation (UP)")
mean(abs(mres_es_up$mean_lfc ))

message("Are mean log2FC values for FDR prioritisation (UP)")
mean(abs(mres_sig_up$mean_lfc ))

message("ES:FDR ratio (UP)")
mean(abs(mres_es_up$mean_lfc )) / mean(abs(mres_sig_up$mean_lfc ))

message("Are mean log2FC values for ES prioritisation (DOWN)")
mean(abs(mres_es_dn$mean_lfc ))

message("Are mean log2FC values for FDR prioritisation (DOWN)")
mean(abs(mres_sig_dn$mean_lfc ))

message("ES:FDR ratio (DOWN)")
mean(abs(mres_es_dn$mean_lfc )) / mean(abs(mres_sig_dn$mean_lfc ))

```

Make a barplot

```{r,mistake4c,fig.height=6,fig.width=6}

par(mar=c(2,20,2,0))

mres_sig_up_vals <- -log(mres_sig_up$pANOVA)
names(mres_sig_up_vals) <- mres_sig_up$set

mres_sig_dn_vals <- -log(mres_sig_dn$pANOVA)
names(mres_sig_dn_vals) <- mres_sig_dn$set

mres_es_up_vals <- mres_es_up$s.dist
names(mres_es_up_vals) <- mres_es_up$set

mres_es_dn_vals <- -1 * mres_es_dn$s.dist
names(mres_es_dn_vals) <- mres_es_dn$set

par(mfrow=c(2,1))

barplot( mres_sig_up_vals,
  horiz=TRUE,las=1,cex.names=0.65,col="gray",
  main="Up Sig (-log10 p-value)",
  xlab="-log10 p-value",
  cex.main=0.9)

barplot( mres_es_up_vals,
  horiz=TRUE,las=1,cex.names=0.65,col="gray",
  main="Up ES",
  xlab="ES",
  cex.main=0.9)

barplot( mres_sig_dn_vals,
  horiz=TRUE,las=1,cex.names=0.65,col="gray",
  main="Down Sig (-log10 p-value)",
  xlab="-log10 p-value",
  cex.main=0.9)

barplot( mres_es_dn_vals,
  horiz=TRUE,las=1,cex.names=0.65,col="gray",
  main="Down ES",
  xlab="-ES",
  cex.main=0.9)

par(mfrow=c(1,1))
par(mar=c(5.1,4.1,4.1,2.1))

```

## Mistake 5 - gene set list size too small or long

ORA is known to have variable results when using gene lists of different length.
I will vary the foreground gene length from 50 to 7000 and see how many significant hits
I get.

```{r,mistake5,fig.height=5,fig.width=7}

up <- rownames(subset(def, log2FoldChange>0 & padj<0.05))
up <- sapply(strsplit(up," "),'[[',1)
up <- unique(gtnew[match(up,gtnew$V1),2])
up <- up[which(!is.na(up))]

dn <- rownames(subset(def, log2FoldChange<0 & padj<0.05))
dn <- sapply(strsplit(dn," "),'[[',1)
dn <- unique(gtnew[match(dn,gtnew$V1),2])
dn <- dn[which(!is.na(dn))]

bg <- rownames(def)
bg <- sapply(strsplit(bg," "),'[[',1)
bg <- unique(gtnew[match(bg,gtnew$V1),2])
bg <- bg[which(!is.na(bg))]

message("Length of up, down and background gene lists. Only for context.")
lapply(list(up,dn,bg),length)

sizes <- c(50,100,200,300,400,500,750,1000,1500,2000,2500,3000,4000,5000,6000,7000)

fup <- nrow(subset(fora(pathways=gsets, genes=up, universe=bg, minSize = 5, maxSize = Inf),padj<0.05))
fdn <- nrow(subset(fora(pathways=gsets, genes=dn, universe=bg, minSize = 5, maxSize = Inf),padj<0.05))

nups <- unlist(lapply(sizes, function(i) {
  ids <- rownames(subset(def2, log2FoldChange>0))
  ids <- head(ids,i)
  genes <- unique(gtnew[match(ids,gtnew$V1),2])
  fres <- fora(pathways=gsets, genes=genes, universe=bg, minSize = 5, maxSize = Inf)
  nrow(subset(fres,padj<0.05))
}))

ndns <- unlist(lapply(sizes, function(i) {
  ids <- rownames(subset(def2, log2FoldChange<0))
  ids <- head(ids,i)
  genes <- unique(gtnew[match(ids,gtnew$V1),2])
  fres <- fora(pathways=gsets, genes=genes, universe=bg, minSize = 5, maxSize = Inf)
  nrow(subset(fres,padj<0.05))
}))

df <- data.frame("up"=nups,"dn"=ndns,row.names=sizes)
df$tot <- df$up + df$dn
df %>%
  kbl(caption="Table XX. Number of differentially regulated gene sets when varying the foreground gene list length.") %>%
  kable_paper("hover", full_width = F)

```

Now to make a line chart of this data.
The first chart is the whole range of data, and the second is a zoom in to 50 to 500 genes.
The line chart shown below corresponds to Figure 4.

```{r,mistake5_linechart,fig.height=5,fig.width=7}

plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",bty="n")
lines(sizes,df$dn,type="b",col="blue")

plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",
  xlim=c(0,500), ylim=c(0,125),bty="n")
lines(sizes,df$dn,type="b",col="blue")

par(mfrow=c(1,2))
plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",bty="n")
lines(sizes,df$dn,type="b",col="blue")

plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",
  xlim=c(0,500), ylim=c(0,125),bty="n")
lines(sizes,df$dn,type="b",col="blue")

unlink("img/5m.png")
png("img/5m.png", width = 700, height = 480)
par(mfrow=c(1,2))
plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",bty="n")
lines(sizes,df$dn,type="b",col="blue")
plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",
  xlim=c(0,500), ylim=c(0,125),bty="n")
lines(sizes,df$dn,type="b",col="blue")
dev.off()
par(mfrow=c(1,1))

unlink("img/5m.pdf")
pdf("img/5m.pdf", width = 7, height = 5)
par(mfrow=c(1,2))
plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",bty="n")
lines(sizes,df$dn,type="b",col="blue")
plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",
  xlim=c(0,500), ylim=c(0,125),bty="n")
lines(sizes,df$dn,type="b",col="blue")
dev.off()
par(mfrow=c(1,1))

```

## Mistake 6 - combining up and downregulated genes into the same test.

Combining up and down into a single test can reduce the number of results significantly.
This is because gene sets show typically have a predominantly pattern of up or down
regulation but not both.
The combined approach discovers only a few unique sets (10 in this case).
The euler plot shown below corresponds to Figure 5.

```{r,mistake6a,fig.height=4,fig.width=5}

message("No. up genes.")
length(up)

message("No. down genes.")
length(dn)

message("No. unique DE genes.")
updn <- unique(c(up,dn))
length(updn)

fres_updn_bg <- fora(pathways=gsets, genes=updn, universe=bg, minSize = 5, maxSize = Inf)
fres_updn_bg <- subset(fres_updn_bg,padj<0.05)$pathway

message("no. significant sets using the combined test:")
length(fres_updn_bg)

message("no. significant upregulated sets using the separate test:")
length(fres_up_bg)

message("no. significant downregulated sets using the separate test:")
length(fres_dn_bg)

message("no. significant sets using the separate tests:")
length(unique(c(fres_up_bg,fres_dn_bg)))

v1 <- list("up"=fres_up_bg,"dn"=fres_dn_bg,"combined"=fres_updn_bg)
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

unlink("img/6m. png")
png("img/6m.png", width = 480, height = 300)
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
dev.off()

unlink("img/6m. pdf")
pdf("img/6m.pdf", width = 5, height = 4)
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
dev.off()

```

## Mistake 7 - using shallow gene annotations

Here, the idea is to show that the gene set database being used makes a huge difference
to the quality of the results.
In particular, the more comprehensive the gene sets, the more comprehensive the
pathway enrichment will be.
This corresponds to Table 3.

```{r,m8a}

msdb <- gmtPathways("ref/msigdb/msigdb.v2025.1.Hs.symbols.gmt")
kegg <- msdb[grep("KEGG",names(msdb))]
kegg_med <- msdb[grep("KEGG_MED",names(msdb))]
kegg <- kegg[grep("KEGG_MED",names(kegg),invert=TRUE)] #separate KEGG LEG from KEGG MED
gobp <- msdb[grep("GOBP",names(msdb))]
reactome <- msdb[grep("REACTOME",names(msdb))]
gslist <- list("KEGG"=kegg,"KEGGM"=kegg_med,"Reactome"=reactome,"GOBP"=gobp,"MSigDB"=msdb)

# no. gene sets
d1 <- unlist( lapply(gslist,length) )

# total number of annotations
d2 <- unlist( lapply(gslist,function(i) { length(unlist(i)) } ) )

# median gene set size
d3 <- unlist(lapply(gslist,function(i) { median(unlist(lapply(i,length))) } ))

# number of genes with one or more functions
d4 <- unlist( lapply(gslist,function(i) { length(unique(unlist(i))) } ))

kegg_up <- fora(pathways=kegg, genes=up, universe=bg, minSize = 5, maxSize = Inf)
kegg_dn <- fora(pathways=kegg, genes=dn, universe=bg, minSize = 5, maxSize = Inf)


message("no. significantly upregulated KEGG pathways")
nrow(subset(kegg_up,padj<0.05))
message("no. significantly downregulated KEGG pathways")
nrow(subset(kegg_dn,padj<0.05))
message("no. significantly dysregulated KEGG pathways")
sum(nrow(subset(kegg_up,padj<0.05)), nrow(subset(kegg_dn,padj<0.05) ))

kegg_med_up <- fora(pathways=kegg_med, genes=up, universe=bg, minSize = 5, maxSize = Inf)
kegg_med_dn <- fora(pathways=kegg_med, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
message("no. significantly upregulated KEGG medicus pathways")
nrow(subset(kegg_med_up,padj<0.05))
message("no. significantly downregulated KEGG medicus pathways")
nrow(subset(kegg_med_dn,padj<0.05))
message("no. significantly dysregulated KEGG medicus pathways")
sum(nrow(subset(kegg_med_up,padj<0.05)), nrow(subset(kegg_med_dn,padj<0.05) ))

reactome_up <- fora(pathways=reactome, genes=up, universe=bg, minSize = 5, maxSize = Inf)
reactome_dn <- fora(pathways=reactome, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
message("no. significantly upregulated Reactome pathways")
nrow(subset(reactome_up,padj<0.05))
message("no. significantly downregulated Reactome pathways")
nrow(subset(reactome_dn,padj<0.05))
message("no. significantly dysregulated Reactome pathways")
sum(nrow(subset(reactome_up,padj<0.05)), nrow(subset(reactome_dn,padj<0.05) ))

gobp_up <- fora(pathways=gobp, genes=up, universe=bg, minSize = 5, maxSize = Inf)
gobp_dn <- fora(pathways=gobp, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
message("no. significantly upregulated GO BP terms")
nrow(subset(gobp_up,padj<0.05))
message("no. significantly upregulated GO BP terms")
nrow(subset(gobp_dn,padj<0.05))
message("no. significantly dysregulated GO BP terms")
sum(nrow(subset(gobp_up,padj<0.05)), nrow(subset(gobp_dn,padj<0.05) ))

msdb_up <- fora(pathways=msdb, genes=up, universe=bg, minSize = 5, maxSize = Inf)
msdb_dn <- fora(pathways=msdb, genes=dn, universe=bg, minSize = 5, maxSize = Inf)

message("no. significantly upregulated MSigDB gene sets")
nrow(subset(msdb_up,padj<0.05))

message("no. significantly downregulated MSigDB gene sets")
nrow(subset(msdb_dn,padj<0.05))

message("no. significantly dysregulated MSigDB gene sets")
sum(nrow(subset(msdb_up,padj<0.05)), nrow(subset(msdb_dn,padj<0.05) ))

up_pw <- c(nrow(subset(kegg_up,padj<0.05)),nrow(subset(kegg_med_up,padj<0.05)),
  nrow(subset(reactome_up,padj<0.05)), nrow(subset(gobp_up,padj<0.05)),
  nrow(subset(msdb_up,padj<0.05)) )

dn_pw <- c(nrow(subset(kegg_dn,padj<0.05)), nrow(subset(kegg_med_dn,padj<0.05)),
  nrow(subset(reactome_dn,padj<0.05)), nrow(subset(gobp_dn,padj<0.05)),
  nrow(subset(msdb_dn,padj<0.05)) )

m8tbl <- data.frame(d1,d2,d3,d4,up_pw, dn_pw)

colnames(m8tbl) <- c("No. gene sets","Total no. annotations","Median gene set size",
  "No. genes with ≥1 annotation","Up-regulated","Down-regulated")

m8tbl %>%
  kbl(caption="Table XX. Gene set metrics, and number of significantly differentially regulated pathways (FDR<0.05).") %>%
  kable_paper("hover", full_width = F)

saveRDS(m8tbl,"m8tbl.Rds")

```

## Mistake 8 - outated gene sets

Here, we want to show how the depth of the gene set library increases over time.
To achieve this, I downloaded all the Reactome gene sets that were available in
MSigDB going back to 2010.
I also included the most recent gene set library from Reactome directly - the
one we used above.
There are a few ways to show the gene set library size including number of gene
sets, mean set size, total number of annotations, and number of unique genes
included.
The charts below show that the number of gene sets increases over time, and
so does the total number of annotations, and the gene coverage.
For simplicity, in the article we just show the number of gene sets.
This corresponds to Figure 6.

```{r,outdated}

par(mar=c(5.1,5.1,4.1,2.1))

reactome_files <- list.files("ref/msigdb/reactome/",full.names=TRUE)
reactome_files <- reactome_files[c(6:20,1:5)]
reactome_files <- c(reactome_files,"ref/ReactomePathways_2025-09-02.gmt")

reactomes <- lapply(reactome_files,gmtPathways)

names(reactomes) <- c("2010_09","2012_09","2013_05","2015_04","2016_01","2016_09","2017_04",
  "2017_10","2018_07","2019_08","2020_03","2020_09","2021_03","2021_04","2022_01","2022_09",
  "2023_03","2023_10","2024_08","2025_06","2025_09")

nogs <- unlist(lapply(reactomes,length))
nogs

# remove sets with 4 or fewer genes
reactomes_filt <- lapply(reactomes,function(gs) {
  l <- lapply(gs,length)
  gs[which(l>=5)]
})
nogs <- unlist(lapply(reactomes_filt,length))
nogs

barplot(nogs,las=2,ylim=c(0,2500),ylab="no. gene sets",main="Reactome gene sets over time",
  border=NA)
lapply(seq(0,2500,500),function(h) { abline(h=h,lty=2,col="gray") })

unlink("img/9m.png")
png("img/9m.png", width = 480, height = 300)
barplot(nogs,las=2,ylim=c(0,2500),ylab="no. gene sets",main="Reactome gene sets over time",
  border=NA,col="darkgray")
lapply(seq(0,2500,500),function(h) { abline(h=h,lty=2,col="lightgray") })
dev.off()

unlink("img/9m.pdf")
pdf("img/9m.pdf", width = 5, height = 4)
barplot(nogs,las=2,ylim=c(0,2500),ylab="no. gene sets",main="Reactome gene sets over time",
  border=NA,col="darkgray")
lapply(seq(0,2500,500),function(h) { abline(h=h,lty=2,col="lightgray") })
dev.off()

reactomes <- reactomes_filt

tot_anno <- unlist(lapply(reactomes,function(gs) {
  length(unlist(gs))
}))
tot_anno

barplot(tot_anno,las=2,ylim=c(0,140000),main="Total no. annotations")

gene_coverage <- unlist(lapply(reactomes,function(gs) {
  length(unique(unlist(gs)))
}))
gene_coverage

barplot(gene_coverage,las=2,ylim=c(0,12000),ylab="gene coverage")

```

## Session information

For reproducibility

<br><details><summary><b>Click HERE to show session info</b></summary><br><p>

```{r,session}

save.image("analysis.Rdata")
sessionInfo()

```

</details>
