---
title: "10 common mistakes that could ruin your enrichment analysis"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: 10mistakes.bib
csl: f1000research.csl
---

<strong>Anusuiya Bora<sup>1,2</sup> & Mark Ziemann<sup>1,2*</sup></strong>

1. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.

2. Burnet Institute, Melbourne, Australia

(*) Correspondence: mark.ziemann@burnet.edu.au

| Author | ORCID |
| --- | --- |
| Anusuiya Bora | 0009-0006-2908-1352 |
| Mark Ziemann | 0000-0002-7688-6974 |

Target journal: F1000 Research

# Abstract

Functional enrichment analysis (FEA) is an incredibly powerful way to summarise complex
genomics data into information about the regulation of biological pathways including cellular
metabolism, signalling and immune responses.
About 10,000 scientific articles describe using FEA each year, making it among the most
used techniques in bioinformatics.
While FEA has become a routine part of workflows via myriad software packages and easy-to-use
websites, mistakes can easily creep in due to poor tool design and unawareness among users of
pitfalls.
Here we outline 10 mistakes that undermine the effectiveness of FEA which we commonly see in
research articles.
We provide practical advice on their mitigation.

# Background

PubMed searches indicate keywords like "pathway analysis" and "enrichment analysis" appear in title
or abstracts of approximately 10,000 articles per year, and that the number of articles matching these
keywords has increased by a factor of 5.4 between 2014 and 2024.
The purpose of FEA is to understand whether gene categories are collectively
differentially represented in the molecular profile at hand, and involves querying hundreds or
thousands of functional categories representing gene pathways or ontologies.
There are a variety of methods for FEA, but the main two are over-representation analysis (ORA) and
functional class scoring (FCS) @Khatri2012-yq.
ORA involves selecting genes based on a hard cut-off followed by a test of enrichment (e.g.: Fisher's
exact test) as compared to a background list @Tavazoie1999-as.
Popular ORA tools include websites like DAVID @Sherman2022-kg and software packages like
ClusterProfiler @Wu2021-mg.
FCS involves ranking all detected genes followed by a test to asses whether the distribution of scores
deviates towards the up- or down-regulated directions.
GSEA @Subramanian2005-bd is a stand-alone FCS software with a graphical user interface, and there
are several command-line implementations such as fgsea @Korotkevich2021-xm.

Recommendations on correct application of pathway enrichment have been previously published
[@Tilford2009-kh; @Tipney2010-sy; @Chicco2022-aa; @Zhao2023-av], yet we and others continue to
observe blatant mistakes and methodological deficiencies appearing in peer-reviewed publications at
an alarming rate [@Timmons2015-yb; @Wijesooriya2022-we].
The purpose of this opinion article is to share what our group has learned about successful FEA over
the past 15 years having authored several articles using it and critically examining hundreds of
published articles using the method.

Using an example RNA-seq dataset and simulation analysis, we provide evidence to show just how
impactful these mistakes are.
Details of the this analysis are provided in the Methods and Supplementary Material.

```{r,setup,echo=FALSE}

#knitr::opts_chunk$set(dev = 'svg')

library("kableExtra")
library("png")
library("grid")

if ( ! file.exists("analysis.html") ) {
  rmarkdown::render("analysis.Rmd")
}

```

# 1. Using uncorrected p-values for statistical significance 

Enrichment tests generate p-values (probability values) between 0 and 1.
The p-value estimates the probability of an observed result occurring by random chance.
A low p-value (eg: p<0.05) suggests the observed result would be unlikely from random data,
suggesting a real effect.
However, as gene set libraries can contain thousands of categories, we could expect 5% of the
gene sets to meet the p<0.05 threshold with random data.
In the example dataset (see Methods below), we randomly selected 1000 genes that met the detection threshold and submitted
these to ORA with gene sets from Reactome.
Of the 1840 Reactomes with five or more members detected, on average we obtained a mean of 51.6 hits (p<0.05; 100 repeats)
with these random genes (**Figure 1A**), proving that reporting raw p-values is bound to yield many
false positives, in this case at a rate of 2.8%.
There are a variety of p-value correction methods to reduce the risk of false positives @Tilford2009-kh,
including approaches from Sidak @Ury01021976, Holm-Bonferroni @holm1979simple and Benjamini-Hochberg
@benjamini1995controlling.
The Benjamini-Hochberg method, also called the false discovery rate (FDR) method, appears to be the
most widely used in genomics to adjust p-values, but it has been critiqued as being overly
conservative when a larger fraction of tests are not null @storey2003statistical.
After applying FDR adjustment to the results of the random gene sets (**Figure 1A**), we see that
the mean number of significant hits (FDR adjusted p<0.05) is 0.16, effectively eliminating false
positives.

In the example dataset (AML3 cells with and without azacitidine treatment), omitting FDR correction leads to the
identification of 696 Reactomes with p<0.05, of which 345 are likely false positives (~49.6%)
(**Figure 1B**).
Omitting p-value correction could lead to half of the enrichment results being false, which is why it
is number one on our list.
Most enrichment analysis packages provide adjusted p-values, so if the tool you're using doesn't, then
it is time for a change.

```{r,1a,fig.cap="Figure 1A. FDR control reduces false positives.",echo=FALSE,fig.height=4,fig.width=4}

mypng = readPNG('img/1a.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

```{r,1b,fig.cap="Figure 1B. Impact of FDR correction of p-values on the number of \'significant\' gene sets.",echo=FALSE,fig.height=3,fig.width=3}

mypng = readPNG('img/1b.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

# 2. Not using a custom background gene list 

Every omics analysis has its limitations.
In the world of gene expression, microarrays can only measure genes it has been designed to assay.
RNA-seq has certain genes that are poorly detected as a result of sequence similarity or GC bias.
Biological differences also play a big part in what is detected @Timmons2015-yb.
Out of the 78,691 human genes annotated in Ensembl's latest release (v115), typically only 12-20k are
expressed at detectable levels with RNA-seq in any one tissue.
In the example dataset based on an earlier Ensembl release (v90) that involves 58,302 annotated genes,
most are silent.
For example, 45,134 recorded a mean less than 10 reads per sample across the six samples.
Only 13,168 genes met this detection threshold.
Although 77.4% of genes are discarded at this step, they account for a miniscule 0.2% of reads.
Filtering poorly detected genes gives a slight boost to differential gene identification
@sha2015effect, but it is most crucial in defining the background gene list for subsequent enrichment
analysis.
Those genes are expressed at such low levels that they wouldn't make a major contribution to the
pathway's activation state.

Using a simulation approach we can demonstrate the consequence of omitting a background on an RNA-seq
experiment.
By drawing a random set of 1000 genes from among the ones with expression above 10 reads per sample
on average in the AML3 example dataset and using an enrichment test (hypergeometric) that uses the
whole genome annotation as a background we see that on average we should expect 444 gene sets
reaching the FDR<0.05 significance level (**Figure 2A**).
Meanwhile, if we use a custom background gene list composed of genes meeting the 10 reads per sample
threshold, we should expect on average 0.16 genesets with FDR<0.05, practically eliminating false
positives.
Many of these false positives are highly reproducible (**Figure 2B**), and as many of them are
related to cancer-relevant cell functions like cell cycle, immune function and signalling, they have the
potential to mislead readers.

Now coming back to the example dataset, we did ORA with a whole genome background and separately with
the custom background of detected genes.
The whole genome background gave a much larger number of "significant" gene sets (1269) in contrast
to the custom background analysis (351) (**Figure 2C**).
The overlap between them was just 347, giving a Jaccard statistic of 0.27.
In other words, without the right background gene list, up to 73% of results could be false positives.

```{r,2a,fig.cap="Figure 2A. Impact of background list on the number of significant gene sets. 100 simulations.",echo=FALSE,fig.height=4,fig.width=4}

mypng = readPNG('img/2a.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

```{r,2b,fig.cap="Figure 2B. Gene sets consistently appearing as false positives include those related to cancer.",echo=FALSE,fig.height=4,fig.width=4}

mypng = readPNG('img/2b.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

```{r,2c,fig.cap="Figure 2C. Impact of background list on the number of significant gene sets. Example dataset. The correct background based on the implementation of detection threshold is labelled 'bg', while the incorrect whole genome background is lebelled 'bg*'.",echo=FALSE,fig.height=3,fig.width=3}

mypng = readPNG('img/2c.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

# 3. Using a tool that doesn’t report enrichment scores 

FDR values can tell us whether something is statically significant, but it doesn’t directly
indicate whether there will be any biological impact [@sullivan2012using;@schober2018statistical].
For that, we need some measure of effect size.
In enrichment analysis, we can use an enrichment score as a proxy measure of effect size.
For rank-based tools like GSEA, the enrichment score varies from -1  to +1 denoting the
distribution of genes in a gene set relative to all other genes.
A score of 1.0 would mean that X genes in the set are the X most upregulated, while if a
value is close to 0, it means the distribution of genes is close to what you might get by
random chance.
For over-representation methods like DAVID, the fold enrichment score is often quoted, which
is the odds ratio of genes of a gene set in the list as compared to the background.
Unfortunately, DAVID doesn’t provide the fold enrichment scores in the main results page,
they are only available in the table for download.
Many other common tools don’t even calculate the enrichment scores (looking at you
clusterProfiler), which leaves researchers in the dark about their effect sizes.
Tools that do provide enrichment scores include ShinyGO (web) @Ge2020-lr, GSEA @Subramanian2005-bd
and fgsea (fora) @Korotkevich2021-xm.

# 4. Prioritising results solely by p-value 

Pathway enrichment analysis can return hundreds of significant results, which can be
confusing to interpret.
Many tools by default will sort the results by significance, but this can lead you to
prioritise pathways that are very large but where each gene is only slightly dysregulated.
To demonstrate this, see the results from a typical pathway enrichment analysis result with
p-value prioritisation and with enrichment score (ES) prioritisation after removal of
non-significant pathways **(Table 1 and Table 2)**.

```{r,m4tbl,echo=FALSE}

m4 <- readRDS("m4tbls.Rds")

mytbl <- m4[[3]]
mytbl <- mytbl[,c(1,2,3,5,4,6)]
colnames(mytbl) <- c("Gene set","Set Size","p-value","FDR","ES","Mean Log2FC")

mytbl |>
  kbl(caption="Table 1. Top upregulated pathways when prioritised by FDR. This emphasises larger and more generic functions.",row.names=FALSE) |>
  kable_paper("hover", full_width = F)

mytbl <- m4[[1]]
mytbl <- mytbl[,c(1,2,3,5,4,6)]
colnames(mytbl) <- c("Gene set","Set Size","p-value","FDR","ES","Mean Log2FC")

mytbl |>
  kbl(caption="Table 2. Top upregulated pathways when prioritised by ES. Pathways with FDR>0.05 were excluded. This emphasises smaller, more specific categories with larger effect sizes.",row.names=FALSE) |>
  kable_paper("hover", full_width = F)

```

P-value prioritisation (**Table 1**) emphasises generic functions with really large gene
sets and moderate fold changes, while enrichment score prioritisation (**Table 2**)
highlights smaller gene sets with highly specific functions, where member genes have bigger
fold changes.
These more specific gene sets are in general better candidates for downstream validation due
to their explanatory power.
The scatterplot shown in **Figure 3** shows how focusing on statistical significance only
could overlook interesting results with larger effect sizes.
End users should therefore use both prioritisation approaches to interpret their data.

```{r,m4fig,fig.cap="Figure 3. Scatterplot showing absolute enrichment scores (x-axis) and log-transformed significance values (y-axis). Gene sets with FDR<0.05 are highlighted in red.",echo=FALSE,fig.height=5,fig.width=5}

mypng = readPNG('img/3a.png')
grid.raster(mypng, x=0.5, y=.5, width=1)

```

# 5. Using gene lists that are too large or too small for ORA 

It’s a common misconception that only differentially expressed genes that meet the FDR
threshold should be submitted to an enrichment test, but this simply isn’t true.
So long as you’re using proper FDR control of your pathways (See #1 above), you can select
genes in any arbitrary way your like.
The caveat here is that enrichment tests (like hypergeometric method) have size ranges of
input gene lists that work best.
We tested different input gene list sizes in ORA and found that 2500 yielded the most (456
with FDR<0.05), while sizes 200 and less yielded very few significant pathways (**Figure 4**).
In the range of 300-1000, there’s a steep increase in the number of significant pathways, and
after that the gradient reduces.
This suggests a sweet spot around 1000, which in our example is 7.6% of the 13,168 genes
detected.
If you want to avoid making arbitrary thresholds (which seem to annoy reviewers) then we’d
suggest using an FCS method like GSEA instead that calculates enrichment from all detected
genes.

```{r,5m,fig.cap="Figure 4. Effect of gene list size on number of significant pathways. Up-regulated in red, down-regulated in blue.",echo=FALSE,fig.height=4.5,fig.width=7}

mypng = readPNG('img/5m.png')
grid.raster(mypng, x=0.5, y=.5, width=1)

```

# 6. Combining up and down-regulated genes in the same ORA test 

In some articles we’ve read, we noticed that authors didn’t conduct separate ORA tests for
up- and down-regulated gene lists, instead opting to submit the combined list for ORA.
This isn’t necessarily an error, as it tests the hypothesis that some pathways are
“dysregulated”, a mix of up- and down-regulated genes, that appear at an elevated rate.
The combined approach can miss a lot of results as compared to the separate approach.
According to the results of our example analysis, the separate approach identified 355
pathways and the combined approach found only 149, that’s 58% fewer (see **Figure 5**).
The combined approach could uniquely identify some pathways, but this is relatively few.
In the example dataset, only 2.2% of results were identified exclusively with the combined
test.

The reason behind this is two-fold.
Firstly, we know that genes in the same pathway are typically correlated with each other
@Gatti2010-tb.
Consider cell cycle genes, or genes responding to pathogens, which are activated in unison
to coordinate a complex biological process.
In a typical differential expression experiment after a stimulus, this results in pathways
that are predominantly up or down regulated, but rarely a mix of up and down.
Due to this phenomenon, the up and down lists each have relatively strong enrichments, but
they are diluted when combined @Hong2014-yn.
Based on this, failing to report data of the separate approach could leave you 58% fewer
results, and an incomplete picture of what's happening at the molecular level.

```{r,6m,fig.cap="Figure 5. Combining up and downregulated genes into one ORA test yields far fewer results.",echo=FALSE,fig.height=3,fig.width=5}

mypng = readPNG('img/6m.png')
grid.raster(mypng, x=0.5, y=.5, width=1)

```

# 7. Using shallow gene annotations 

One of the most important decisions you’ll make when doing a pathway enrichment analysis is
selecting the database to query.
There are many options to consider, both proprietary and open source.
When choosing, consider whether the database contains the gene sets that you a priori
suspect will be altered in the profile you’re looking at.
Secondly consider the breadth and depth of the pathway library, this will be where the
unexpected discoveries may occur, and it pays to use a comprehensive one to capture as many
aspects of your data as possible.
To demonstrate this, see how KEGG legacy and KEGG Medicus seem tiny when compared to Reactome,
which is itself dwarfed by Gene Ontology’s Biological Process (GOBP; Table 3).
Consequently, the results obtained for the example dataset are substantially richer for
Reactome and GOBP as compared to KEGG libraries.

```{r,m8tbl,echo=FALSE}

m8 <- readRDS("m8tbl.Rds")
m8 |>
  kbl(caption="Table 3. Size metrics of selected gene set libraries, and the number of differentially regulated pathways in the AML3 experiment (FDR<0.05).") |>
  kable_paper("hover", full_width = F)

```

# 8. Using outdated gene identifiers and gene sets 

One great thing about working with omics data is that data repositories like GEO @clough2024ncbi
are loaded with previously published data sets that we can reanalyse with new pathway
databases and software tools to eke out further insights.
When the data is several years old, we should use the processed data with caution, as many
gene names may have since changes.
For example, as previously demonstrated, Illumina’s EPIC DNA Methylation microarray was
released in 2016, and in the following eight years, 3,253 of 22,588 gene names on the chip
have changed (14.4%) @Ziemann2024-be, meaning that those genes wouldn’t be recognised by
the pathway enrichment software.
To update defunct gene symbols, the HGNChelper R package can help @Oh2020-ku, and it
has the benefit of fixing gene symbols ruined by Excel autocorrect, which are unfortunately
common in GEO @Ziemann2016-hp.
Persistent gene identifiers like Ensembl (eg: ENSG00000180096) and HGNC (eg: HGNC:2879)
are less likely to change over time and would therefore be preferable over gene symbols
(eg: *SEPTIN1*) for FEA.

The depth of pathway databases increases every year as annotation consortia continue
assimilating functional information from the literature, and this impacts the
quality of results and the conclusions that can be derived @wadi2016impact.
Sometimes these databases undergo large updates, as shown by the chart of Reactome growth below
(**Figure 6**).
To be certain you have the best possible gene annotation for your analysis, it’s always best
to download the newest version.

```{r,9m,fig.cap="Figure 6. Reactome gene set growth over time. Gene sets were downloaded from the MSigDB website, except The last bar which represents the latest gene sets downloaded directly from Reactome, not yet incorporated into MSigDB.",echo=FALSE,fig.height=3,fig.width=5}

mypng = readPNG('img/9m.png')
grid.raster(mypng, x=0.5, y=.5, width=1)

```

# 9. Bad presentation

When discussing FEA with colleagues, it is often said that the method is used to generate
"filler"; data and charts used to "pad out" articles that would otherwise be too short or
lack the normal number of figures.
For FEA, a general rule of thumb is to only show the charts and data that are relevant to
assessing aims and hypotheses, and contribute to the conclusions.
While others have recommended using multiple FEA tools @Chicco2022-aa, we'd suggest limiting
the amount of data shown in an article to just one FEA method and one or two gene set
databases (eg: Reactome, transcription factor target genes).
Excessive use of tools and databases can make the results difficult to interpret, as in
@lin2019indoxyl.

We've also noticed cases of outright confusing, incomplete and simply wrong data
presentatation choices that you should avoid:

1. The number of selected genes in category is often shown as evidence of enrichment,
[@zhao2019itraq; @li2019candidate;@sariman2019investigation;@bhatia2019long],
but this can be misleading because this is one of four numbers that goes into calculating
a fold enrichment score.

2. Similarly, the proportion of selected genes that belong to a gene category is sometimes
shown [@wang2019osteoporosatlas;@wang2019comparison;@hu2019itraq;@koh2019baicalein],
but this does not directly reflect the fold enrichment score.

3. Presenting enrichment results as a pie chart
[@liu20196;@li2019candidate;@hu2019itraq;@lin2019indoxyl]
isn't recommended because it isn't possible to show enrichment scores and significance
values in this form.

4. Sometimes a network of genes or pathways are shown, but the significance of nodes and
edges aren't described @bandi2019hepatic.

5. Figures missing key elements such as axis labels
[@boyko2019reference;@lou2019high;@shi2019physiological;@li2019identification;
@li2019candidate].

6. FEA mentioned in the abstract but no data shown in the main article
[@xu2019sirt2; @di2019genome].

7. Confusion around which tool was used for each figure and panel (eg: @jin2019expression).

# 10. Neglecting methods reproducibility

According to Goodman and colleagues @goodman2016does, methods reproducibility is:

> "the provision of enough detail about study procedures and data so the same procedures
could, in theory or in actuality, be exactly repeated."

There are several crucial pieces of information required in order to enable reproducibility
of enrichment analysis including;

* how genes were selected or scored,

* the tool used, and it's version,

* the options or parameters applied,

* the statistical test used,

* the gene set/pathway database(s) queries, and their versions,

* for ORA, how a background list was defined, and,

* how p-value correction was done [@Chicco2022-aa;@wijesooriya2021guidelines].

A systematic literature analysis published in 2022 found insufficient background description
in 95% of articles describing ORA tests, and p-value correction was insufficiently
described in 57% of articles, suggesting that enrichment analysis generally suffers from an
atrocious lack of methods reproducibility @Wijesooriya2022-we.
Indeed, this literature analysis turned up some memorably poor methodological descriptions
including:

>"GO and KEGG enrichment analysis were performed on the DEGs by perl scripts in house."
@zhou2019transcriptome.

>"Gene ontology (GO) and pathway analysis were performed in the standard enrichment
computation method." @liu2019long.

We've also noted some cases where FEA wasn't described in the methods at all, despite being important
enough to mention said results in the abstract [@hu2019high; @zhao2019characterization;
@chen2019usf1].
Moreover, we've identified cases where the tool mentioned in the methods section is inconsistent
with what's shown in the results [@li2019spag5;@tong2019combined].

In addition to including the methodological details mentioned above, authors could also
provide gene profile data and/or gene lists used for enrichment analysis as supplementary
files, or better still, providing full reproducibility and transparancy with the five
pillars framework @ziemann2023five.

# Other issues

There are several more subtle issues not covered in depth here but are worth mentioning because they
have been flagged as potential problems.
First, the length of genes is known to impact the ease at which they are detected and so correction of
gene length has been suggested to improve enrichment results [@mi2012length;@mandelboum2019recurrent].
Second, many FEA tests use genes as the sampling unit and do not take into consideration or model biological
variation which could yield inrealistic significance values @goeman2007analyzing.
Third, the size of gene sets, even though they represent similar biology, can disproportionately impact significance
scores and complicate interpretation [@karp2021pathway;@mubeen2022influence].
Fourth, tight correlation between genes' expression within a pathway could exacerbate false positive rates
[@gatti2010heading; @wu2012camera].
Fifth, slight differences in the implementation of ORA tests can impact results in some circumstances
@ziemann2024two.
Lastly, some web-based FEA tools lack longevity.
For example, DAVID 6.8 [@huang2009systematic;@Sherman2022-kg] has been used for over 10,000 publications but has
since 2022 been taken offline leaving those articles in an irreproducible state.
As web-based tools appear to be most popular option for FEA [@xie2021popularity;@Wijesooriya2022-we],
we should advocate using tools which allow preservation/archiving of the tool as a Docker image (eg: @Ge2020-lr),
which could enable reproducibility into the future @perkel2020challenge.

# Conclusion

So why are problems so pervasive in enrichment analysis?
It's likely a combination of poor researcher training, supervision and peer-review scrutiny.
The design of tools and (poor) quality of tool documentation might also play a role.
We also know that poor methods have a type of advantage compared to the more rigorous ones
due to researcher preferences to present "significant" findings @smaldino2016natural and reliance upon default settings
even if they are incorrect [@mubeen2022influence;@Wijesooriya2022-we].
Problems 2, 3, 5 and 6 appear to be specific to ORA-based tools, and can be avoided entirely by switching to
FCS tools like GSEA; this has the added benefit of enhanced accuracy in terms of precision and
recall [@kaspi2020mitch;@ziemann2024two].
Although learning and running FCS tools is more difficult and time-consuming, the benefits to the quality of results
are substantial.

The deluge of manuscripts in genomics and other fields places increasing burden on a limited pool of competant,
voluntary peer-reviewers to a point where editors are struggling to maintain the peer-review system as we know it
@adam2025peer.
As studies become ever more multi-disciplinary, this makes it more difficult for editors to cover all the technical
aspects thoroughly, causing mistakes like the above to become widespread.
Ultimately, responsibility for using best practices in enrichment analysis lies with authors.
Avoiding these mistakes will have a number of benefits including avoiding wasted resources on unreliable research directions.

# Methods

The example RNA-seq dataset used involves AML3 cells with and without azacytidine treatment
@lund2014dnmt.
Read counts were obtained from the DEE2 database using the getDEE2 R package using SRA
project accession SRP038101 as a query @ziemann2019digital.
Genes with average expression above 10 counts per sample were considered detected, and
genes not meeting this criterion were removed from downstream analysis (unless stated).
Differential expression was conducted using DESeq2 v1.48.2 @love2014differential.
Human gene symbols were updated to Ensembl 115 @dyer2025ensembl.
Genes with FDR<0.05 were considered significantly differentially expressed.
Reactome gene sets were downloaded in GMT format on 02/09/2025 @milacic2024reactome, and
these gene sets were used for all subsequent enrichment tests unless otherwise stated.
All ORA tests were conducted using the `fora` function belonging to the `fgsea` package
v1.34.2 @korotkevich2016fast.
Minimum gene set size was set to 5 for all subsequent enrichment tests.
Gene sets with FDR<0.05 were considered significantly differentially expressed.

To demonstrate the importance of FDR control (Mistake 1), random sampling of 1000 detected
genes was followed with ORA with and without FDR control.
This was repeated 100 times.
For comparison of the nominal and FDR corrected results, the up (1672) and down (1926)
regulated genes underwent ORA using a background consisting of the 13168 detected genes.

To demonstrate the importance of a suitable background gene list (Mistake 2), random sampling
of 1000 detected genes as the foreground followed by ORA using either the whole genome
background or background consisting of detected genes.
This process was repeated 1000 times.

To demonstrate the importance of using enrichment scores for interpretation of FEA results,
(Mistake 3 and 4) the `mitch` package v1.20.0 was used for FCS analysis of the differential
expression data @kaspi2020mitch.

To investigate how foreground gene set size impact the number of ORA results (Mistake 5),
the top N significant up- and down-regulated genes were selected for ORA compared to the
background consisting of detected genes, where N was varied between 50 and 7000.

To demonstrate the effect of combining up and down regulated genes into a single test
(Mistake 6), we took all unique genes with FDR<0.05 (n=3950) as the foreground for an ORA
test, and compared these results to the separate approach (n=1667,1923 up and
downregulated genes respectively).

To show the effect of using shallow gene annotations (Mistake 7), various gene set
libraries were extracted from the MSigDB Collection (msigdb.v2025.1.Hs.symbols.gmt)
@liberzon2015molecular.
These gene set libraries were each used for ORA tests.

To show the effect of using old gene sets (Mistake 8), the Reactome gene sets were extracted
from archived MSigDB Collections going back to 2010.
These were compared to the most recent Reactome release (02/09/2025).

Instances of poor presentation (Mistake 9) and methods reproducibility (Mistake 10)
were drawn from unpublished notes made during a previous systematic analysis of the
literature @Wijesooriya2022-we.

Analysis was conducted in R 4.5.1 using a Bioconductor Docker image corresponding to
Bioconductor release 3.21.
Analysis code is available from GitHub (https://github.com/markziemann/10mistakes)
and the Docker image is available from DockerHub (https://hub.docker.com/repository/docker/mziemann/10mistakes/general).
These will be archived to Zenodo upon acceptance.

# Bibliography
