---
title: "10 common mistakes that could ruin your enrichment analysis"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: 10mistakes.bib
csl: briefings-in-bioinformatics.csl
---

<strong>Anusuiya Bora<sup>1,2</sup>, Matthew McKenzie<sup>1</sup> & Mark Ziemann<sup>1,2*</sup></strong>

1. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.

2. Burnet Institute, Melbourne, Australia

(*) Correspondence: mark.ziemann@burnet.edu.au

| Author | ORCID |
| --- | --- |
| Anusuiya Bora | 0009-0006-2908-1352 |
| Matthew McKenzie | 0000-0001-7508-1800 |
| Mark Ziemann | 0000-0002-7688-6974 |

Target journal: Nature Methods

# Abstract

Functional enrichment analysis (FEA) is an incredibly powerful way to summarise complex
genomics data into information about the regulation of biological pathways including cellular
metabolism, signalling and immune responses.
About 10,000 scientific articles describe using FEA each year, making it among the most
used techniques in bioinformatics.
While FEA has become a routine part of workflows via myriad software packages and easy-to-use
websites, mistakes can easily creep in due to poor tool design and unawareness among users of
pitfalls.
Here we outline 10 mistakes that undermine the effectiveness of FEA which we commonly see in
research articles.
We provide practical advice on their mitigation.

# Background

PubMed searches indicate keywords like "pathway analysis" and "enrichment analysis" appear in
titles or abstracts of approximately 10,000 articles per year, and that the number of articles
matching these keywords has increased by a factor of 5.4 between 2014 and 2024.
The purpose of FEA is to understand whether gene categories are collectively
differentially represented in the molecular profile at hand, and involves querying hundreds or
thousands of functional categories representing gene pathways or ontologies.
There are a variety of methods for FEA, but the main two are over-representation analysis
(ORA) and functional class scoring (FCS) @Khatri2012-yq.
ORA involves selecting genes based on a hard cut-off followed by a test of enrichment (e.g.:
Fisher's exact test) as compared to a background list @Tavazoie1999-as.
Popular ORA tools include websites like DAVID @Sherman2022-kg and software packages like
ClusterProfiler @Wu2021-mg.
FCS involves ranking all detected genes followed by a test to assess whether the distribution
of scores deviates towards the up- or down-regulated directions.
GSEA @Subramanian2005-bd is a stand-alone FCS software with a graphical user interface, and
there are several command-line implementations such as fgsea @Korotkevich2021-xm.

Recommendations on correct application of pathway enrichment have been previously published
[@Tilford2009-kh; @Tipney2010-sy; @Chicco2022-aa; @Zhao2023-av], yet we and others continue to
observe blatant mistakes and methodological deficiencies appearing in peer-reviewed
publications at an alarming rate [@Timmons2015-yb; @Wijesooriya2022-we].
Such methodological and interpretation errors are known to "snowball" - a process where an
initial flawed research study can create cascading problems for scientific reliability
@liu2023misuse.

The purpose of this opinion article is to share what our group has learned about successful
FEA over the past 15 years having authored several articles using it and critically examining
hundreds of published articles describing the use of FEA.

Using an example RNA-seq dataset and simulation analysis, we provide evidence to show just how
impactful these mistakes are.
Details of this analysis are provided in the Methods and Supplementary Material.

```{r,setup,echo=FALSE}

#knitr::opts_chunk$set(dev = 'svg')

library("kableExtra")
library("png")
library("grid")

if ( ! file.exists("analysis.html") ) {
  rmarkdown::render("analysis.Rmd")
}

```

# 1. Using uncorrected p-values for statistical significance 

Enrichment tests generate p-values (probability values) between 0 and 1.
The p-value estimates the probability of an observed result occurring by random chance.
A low p-value (eg: p<0.05) suggests the observed result would be unlikely from random data,
suggesting a real effect.
However, as gene set libraries can contain thousands of categories, we could expect 5% of the
gene sets to meet the p<0.05 threshold with random data.
Therefore, we almost always get many "significant" results just by chance.
This leads to the multiple testing problem whereby uncorrected p-values lead to high rates of
false positives.

In the example dataset (see Methods below), we randomly selected 1000 genes that met the
detection threshold and submitted these to ORA with gene sets from Reactome.
Of the 1840 Reactomes with five or more members detected, on average we obtained a mean of
51.6 hits (p<0.05; 100 repeats) with these random genes (**Figure 1A**), proving that
reporting raw p-values is bound to yield many false positives, in this case at a rate of 2.8%.
There are a variety of p-value correction methods to reduce the risk of false positives
@Tilford2009-kh, including approaches from Sidak @Ury01021976, Holm-Bonferroni @holm1979simple
and Benjamini-Hochberg @benjamini1995controlling.
The Benjamini-Hochberg method, also called the false discovery rate (FDR) method, appears to
be the most widely used in genomics to adjust p-values, but it has been critiqued as being
overly conservative when a larger fraction of tests are not null @storey2003statistical.
After applying FDR adjustment to the results of the random gene sets (**Figure 1A**), the
mean number of significant hits (FDR adjusted p<0.05) is 0.16, effectively eliminating false
positives.

In the example dataset (AML3 cells with and without azacitidine treatment), there were 3598
differentially expressed genes, with 1672 up- and 1926 down-regulated.
The top 1000 up- and down-regulated genes by p-value were selected for separate ORA tests.
Omitting FDR correction led to the identification of 627 Reactomes with p<0.05, of which 317
are likely false positives (50.6%) (**Figure 1B**).
Omitting p-value correction could lead to half of the enrichment results being false, which
is why it is number one on our list.
Most enrichment analysis packages provide adjusted p-values, so if the tool you're using
doesn't, then it is time for a change.

```{r,1a,fig.cap="Figure 1A. FDR control reduces false positives.",echo=FALSE,fig.height=4,fig.width=4}

mypng = readPNG('img/1a.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

```{r,1b,fig.cap="Figure 1B. Impact of FDR correction of p-values on the number of \'significant\' gene sets.",echo=FALSE,fig.height=3,fig.width=3}

mypng = readPNG('img/1b.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

# 2. Not using a custom background gene list 

Every omics analysis has its limitations.
In the world of gene expression, microarrays can only measure genes it has been designed to assay.
RNA-seq has certain genes that are poorly detected as a result of sequence similarity or GC bias.
Biological differences also play a big part in what is detected @Timmons2015-yb.
Out of the 78,691 human genes annotated in Ensembl's latest release (v115), typically only 12-20k are
expressed at detectable levels with RNA-seq in any one tissue.
In the example dataset based on an earlier Ensembl release (v90) that involves 58,302 annotated genes,
most are silent.
For example, 45,134 recorded a mean <10 reads per sample across the six samples, with
only 13,168 genes meeting this detection threshold.
Although 77.4% of genes are discarded at this step, they account for a miniscule 0.2% of reads.
Filtering poorly detected genes gives a slight boost to differential gene identification
@sha2015effect, but it is most crucial in defining the background gene list for subsequent enrichment
analysis.
Genes with very low expression are unlikely to be biologically relevant in the context being studied
and should be excluded from the background.

Using a simulation approach, we can demonstrate the consequence of omitting a custom background gene
list on an RNA-seq experiment.
By drawing a random set of 1000 genes in the AML3 example dataset with expression above 10 reads per
sample on average and using an enrichment test (hypergeometric) that uses the whole genome
annotation as a background, we see 444 gene sets reaching the FDR<0.05 significance level on average
(**Figure 2A**).
Alternatively, if we use a custom background gene list composed of genes meeting the 10 reads per
sample threshold, we expect 0.16 gene sets with FDR<0.05, practically eliminating false positives.
These false positives are highly reproducible (**Figure 2B**), and as many of them are
cancer-relevant pathways like cell cycle and immune signalling, they have the
potential to mislead readers.

Performing ORA with azacitidine-responsive genes with the whole genome background gave a much
larger number of "significant" gene sets (1047) in contrast to the custom background analysis
(310) (**Figure 2C**), giving a Jaccard statistic of 0.30.
In other words, without the right background gene list, up to 70% of results could be false
positives.

```{r,2a,fig.cap="Figure 2A. Impact of background list on the number of significant gene sets. 100 simulations.",echo=FALSE,fig.height=4,fig.width=4}

mypng = readPNG('img/2a.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

```{r,2b,fig.cap="Figure 2B. Gene sets appearing as false positives in 100/100 simulations include those related to cancer.",echo=FALSE,fig.height=4,fig.width=4}

mypng = readPNG('img/2b.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

```{r,2c,fig.cap="Figure 2C. Impact of background list on the number of significant gene sets. Example dataset. The correct background is generated by setting a detection threshold specific to the experiment. The incorrect background involves assuming all genes in the annotation are expressed and able to be detected.",echo=FALSE,fig.height=3,fig.width=3}

mypng = readPNG('img/2c.png')
grid.raster(mypng, x=0.5, y=0.5, width=1)

```

# 3. Using a tool that doesn’t report enrichment scores 

FDR values can tell us whether something is statistically significant, but it doesn’t directly
indicate whether there will be any biological impact [@sullivan2012using;@schober2018statistical].
For that, we need some measure of effect size.
In enrichment analysis, we can use an enrichment score as a proxy measure of effect size.
For rank-based tools like GSEA, the enrichment score varies from -1  to +1, denoting the
distribution of genes in a gene set relative to all other genes @Subramanian2005-bd.
For a gene set composed of 15 genes, a score of 1.0 would mean that these 15 genes are the
top 15 upregulated, while if a value is close to 0, it means the distribution of genes is
close to what you might get by random chance.
For over-representation methods like DAVID, the fold enrichment score is often quoted, which
is the odds ratio of genes of a gene set in the foreground list as compared to the background @Tilford2009-kh.
Unfortunately, DAVID doesn’t provide the fold enrichment scores in the main results page,
they are only available in the table for download.
Many other common tools don’t provide enrichment scores (for example clusterProfiler), which
leaves researchers with no information about their effect sizes.
Tools that do provide enrichment scores include ShinyGO (web) @Ge2020-lr, GSEA @Subramanian2005-bd
and fgsea (fora) @Korotkevich2021-xm.

# 4. Prioritising results solely by p-value 

Pathway enrichment analysis can return hundreds of significant results, which can be
confusing to interpret.
Many tools by default will sort the results by significance, but this can lead users to
prioritise pathways that are very large but where each gene is only slightly dysregulated.
To demonstrate this, see the results from a typical pathway enrichment analysis result with
p-value prioritisation and with enrichment score (ES) prioritisation after removal of
non-significant pathways **(Table 1 and Table 2)**.

```{r,m4tbl,echo=FALSE}

m4 <- readRDS("m4tbls.Rds")

mytbl <- m4[[3]]
mytbl <- mytbl[,c(1,2,3,5,4,6)]
colnames(mytbl) <- c("Gene set","Set Size","p-value","FDR","ES","Mean Log2FC")

mytbl |>
  kbl(caption="Table 1. Top upregulated pathways when prioritised by FDR. This emphasises larger and more generic functions.",row.names=FALSE) |>
  kable_paper("hover", full_width = F)

mytbl <- m4[[1]]
mytbl <- mytbl[,c(1,2,3,5,4,6)]
colnames(mytbl) <- c("Gene set","Set Size","p-value","FDR","ES","Mean Log2FC")

mytbl |>
  kbl(caption="Table 2. Top upregulated pathways when prioritised by ES. Pathways with FDR>0.05 were excluded. This emphasises smaller, more specific categories with larger effect sizes.",row.names=FALSE) |>
  kable_paper("hover", full_width = F)

```

P-value prioritisation (**Table 1**) emphasises generic functions with large gene
sets and moderate fold changes, while enrichment score prioritisation (**Table 2**)
highlights smaller gene sets with highly specific functions, where member genes have bigger
fold changes.
These more specific gene sets are (generally) better candidates for downstream validation due
to their explanatory power.
The scatterplot shown in **Figure 3** shows how focusing on statistical significance only
could overlook interesting results with larger effect sizes.
End users should therefore use both prioritisation approaches to interpret their data.

```{r,m4fig,fig.cap="Figure 3. Scatterplot showing absolute enrichment scores (x-axis) and log-transformed significance values (y-axis) for each detected pathway. Gene sets with FDR<0.05 are highlighted in red.",echo=FALSE,fig.height=5,fig.width=5}

mypng = readPNG('img/3a.png')
grid.raster(mypng, x=0.5, y=.5, width=1)

```

# 5. Using gene lists that are too large or too small for ORA 

It’s a common misconception that only differentially expressed genes that meet the FDR
threshold should be submitted to an enrichment test.
Tarca and colleagues suggest a heuristic that selects the top 1% of genes if there are none
that meet the standard significance cut-off @tarca2013comparison.
If proper FDR control of enrichment results is applied (See #1 above), then gene selection
criteria can be flexible.
The caveat is that enrichment tests (like the hypergeometric method) have size ranges of
input gene lists that work best.
We tested different input gene list sizes in ORA and found that 2500 yielded the most (455
with FDR<0.05), while lists of 200 and fewer yielded very few (**Figure 4**).
In the range of 300-1000, there’s a steep increase in the number of significant pathways,
with the gradient reducing with more than 1000 genes.
This result indicates that using a larger gene list of up to 2500 genes would be better, but
the main downside is that these additional statistically significant results correspond to
relatively small enrichment scores.
For example, significantly downregulated pathways at a gene set size of 100, mean fold
enrichment scores were 16.9, and at a size of 2500, they were 2.5.
Users may wish to prespecify a fold enrichment score that they consider biologically relevant
(3-5 seems reasonable), and then tune their analysis to capture the most statistically
significant enrichments above this score.
In the example dataset, the number of gene sets meeting a minimum fold enrichment score
of 3 appeared to peak with a gene set size of 700 for the upregulated genes, and 800 for
downregulated genes (**Figure 4**), which corresponds to 5-6% of all detected genes.

This suggests that a gene list size of 700-800 genes, or 5-6% of all those detected would
be reasonable for a differential expression study.
Nevertheless, some users may want to avoid setting seemingly arbitrary thresholds - in that
case, using an FCS method like GSEA instead that calculates enrichment from all detected
genes would be recommended.

```{r,5m,fig.cap="Figure 4. Effect of gene list size on number of significant pathways. Up-regulated in red, down-regulated in blue.",echo=FALSE,fig.height=4.5,fig.width=7}

mypng = readPNG('img/5m.png')
grid.raster(mypng, x=0.5, y=.5, width=1)

```

# 6. Combining up and down-regulated genes in the same ORA test 

In some articles, we noticed that authors didn’t conduct separate ORA tests for
up- and down-regulated gene lists, instead opting to submit the combined list for ORA.
This isn’t necessarily an error, as it tests the hypothesis that some pathways are
“dysregulated”; a mix of up- and down-regulated genes which appear at an elevated rate.
However, this combined approach can miss many enriched gene sets when compared to the
separate approach.
According to the results of our example analysis, the separate approach identified 289
pathways compared to the combined approach that found only 53 (84% fewer; **Figure 5**).
The combined approach does uniquely identify some pathways, but this is relatively few.
In the example dataset, only 2.0% (6) of gene sets were identified exclusively with the combined
test.

The reason behind this is two-fold.
Firstly, we know that genes in the same pathway are typically correlated with each other
@Gatti2010-tb.
Consider cell cycle genes, or genes responding to pathogens, which are activated in unison
to coordinate a complex biological process.
In a typical differential expression experiment after a stimulus, this results in pathways
that are predominantly up or down regulated, but rarely a mix of up and down.
Due to this phenomenon, the up and down lists each have relatively strong enrichments, but
they are diluted when combined @Hong2014-yn.
Based on this, failing to report data using the separate approach could result in fewer
identified gene sets, and an incomplete picture of what's happening at the molecular
level.

```{r,6m,fig.cap="Figure 5. Combining up and downregulated genes into one ORA test yields far fewer results.",echo=FALSE,fig.height=3,fig.width=5}

mypng = readPNG('img/6m.png')
grid.raster(mypng, x=0.5, y=.5, width=1)

```

# 7. Using shallow gene annotations 

One of the most important decisions for pathway enrichment analysis is selecting the
database to query.
There are many options to consider, both proprietary and open source.
When choosing, users should consider whether the database contains the gene sets that they
a priori suspect will be altered.
Secondly, consider the breadth and depth of the pathway library; this will be where the
unexpected discoveries may occur and it pays to use a comprehensive library to capture as many
aspects of the dataset as possible.
To demonstrate this, note how KEGG legacy and KEGG Medicus seem tiny when compared to Reactome,
which is itself dwarfed by Gene Ontology’s Biological Process (GOBP; Table 3).
Consequently, the results obtained for the example dataset are substantially richer for
Reactome and GOBP as compared to KEGG libraries.

```{r,m8tbl,echo=FALSE}

m8 <- readRDS("m8tbl.Rds")
m8 |>
  kbl(caption="Table 3. Size metrics of selected gene set libraries, and the number of differentially regulated pathways in the AML3 experiment (FDR<0.05).") |>
  kable_paper("hover", full_width = F)

```

# 8. Using outdated gene identifiers and gene sets 

Data repositories like GEO @clough2024ncbi contain thousands of previously published data
sets that we can reanalyse with new pathway databases and software tools to gain further
insights.
However, when the data is several years old, we should use it with caution, as many
gene names may have changed.
For example, Illumina’s EPIC DNA Methylation microarray was released in 2016, and in the
following eight years, 3,253 of 22,588 gene names on the chip have changed (14.4%)
@Ziemann2024-be.
Therefore, these genes wouldn’t be recognised by the pathway enrichment software.
To update defunct gene symbols, the HGNChelper R package can help @Oh2020-ku, also having
the benefit of fixing gene symbols corrupted by Excel autocorrect, which are unfortunately
common in GEO @Ziemann2016-hp.
Persistent gene identifiers like Ensembl (eg: ENSG00000180096) and HGNC (eg: HGNC:2879)
are less likely to change over time and are therefore preferable over gene symbols
(eg: *SEPTIN1*) for FEA.

The depth of pathway databases increases every year as annotation consortia continue
assimilating functional information from the literature, and this impacts the
quality of results and the conclusions that can be derived @wadi2016impact.
These databases can undergo large updates, as shown by the chart of Reactome growth over time
(**Figure 6**).
Therefore, it’s always best to download and use the newest available version of the gene sets.

```{r,9m,fig.cap="Figure 6. The number of Reactome gene sets grows over time. Gene sets were downloaded from the MSigDB website, except The last bar which represents the latest gene sets downloaded directly from Reactome, not yet incorporated into MSigDB.",echo=FALSE,fig.height=3,fig.width=5}

mypng = readPNG('img/9m.png')
grid.raster(mypng, x=0.5, y=.5, width=1)

```

# 9. Bad presentation

When discussing FEA with colleagues, it is often said that the method is used to generate
"filler"; data and charts used to "pad out" articles that would otherwise be too short or
lack the normal number of figures.
For FEA, a general rule of thumb is to only show the charts and data that are relevant to
assessing aims and hypotheses, and which contribute to the conclusions.
While others have recommended using multiple FEA tools @Chicco2022-aa, we'd suggest limiting
the amount of data shown in an article to just one FEA method and one or two gene set
databases (eg: Reactome, transcription factor target genes).
Excessive use of tools and databases can make the results difficult to interpret, as in
@lin2019indoxyl.

We've also noticed cases of confusing, incomplete or incorrect data
presentation choices that should be avoided:

1. The number of selected genes in a category is often shown as evidence of enrichment,
[@zhao2019itraq; @li2019candidate;@sariman2019investigation;@bhatia2019long],
but this can be misleading because this is one of four numbers that goes into calculating
a fold enrichment score.

2. Similarly, the proportion of selected genes that belong to a gene category is sometimes
shown [@wang2019osteoporosatlas;@wang2019comparison;@hu2019itraq;@koh2019baicalein],
but this does not directly reflect the fold enrichment score.

3. Presenting enrichment results as a pie chart
[@liu20196;@li2019candidate;@hu2019itraq;@lin2019indoxyl]
isn't recommended because it isn't possible to show enrichment scores and significance
values in this form.

4. Sometimes a network of genes or pathways are shown, but the significance of nodes and
edges aren't described @bandi2019hepatic.

5. Figures missing key elements such as axis labels
[@boyko2019reference;@lou2019high;@shi2019physiological;@li2019identification;
@li2019candidate].

6. FEA mentioned in the abstract but no data shown in the main article or supplement
[@xu2019sirt2; @di2019genome].

7. Confusion around which tool was used for each figure and panel [eg: @jin2019expression].

Such misinterpretation and data presentation problems can also occur when a tool is used without
understanding the statistical basis of inference @liu2023misuse, so it is crucial that users
take the time to familiarise themselves with the tool's documentation and recommendations.

# 10. Neglecting methods reproducibility

According to Goodman and colleagues @goodman2016does, methods reproducibility is:

> "the provision of enough detail about study procedures and data so the same procedures
could, in theory or in actuality, be exactly repeated."

There are several crucial pieces of information required in order to enable reproducibility
of enrichment analysis including;

* how genes were selected or scored - especially whether up or down-regulated genes were considered
separately or combined,

* the tool used, and its version,

* the options or parameters applied,

* the statistical test used,

* the gene set/pathway database(s) queries, and their versions,

* for ORA, how a background list was defined, and,

* how p-value correction was done [@Chicco2022-aa;@wijesooriya2021guidelines].

A systematic literature analysis published in 2022 found insufficient background description
in 95% of articles describing ORA tests, and p-value correction was insufficiently
described in 57% of articles, suggesting that enrichment analysis generally suffers from a
lack of methods reproducibility @Wijesooriya2022-we.
Indeed, this literature analysis identified some inadequate methodological descriptions
including:

>"GO and KEGG enrichment analysis were performed on the DEGs by perl scripts in house."
@zhou2019transcriptome.

>"Gene ontology (GO) and pathway analysis were performed in the standard enrichment
computation method." @liu2019long.

We've also noted some cases where FEA wasn't described in the methods at all, despite being important
enough to mention said results in the abstract [@hu2019high; @zhao2019characterization;
@chen2019usf1].
Moreover, we've identified cases where the tool mentioned in the methods section is inconsistent
with what's shown in the results [@li2019spag5;@tong2019combined].

In addition to including the methodological details mentioned above, authors could also
provide gene profile data and/or gene lists used for enrichment analysis as supplementary
files, or better still, provide full reproducibility and transparency with the five
pillars framework @ziemann2023five.

# Other issues

There are several more subtle issues not covered in depth here but are worth mentioning as they
have been flagged as potential problems.
First, the length of genes is known to impact the ease at which they are detected and so correction of
gene length has been suggested to improve enrichment results [@mi2012length;@mandelboum2019recurrent].
Second, many FEA tests use genes as the sampling unit and do not take into consideration (or model) biological
variation which could yield unrealistic significance values @goeman2007analyzing.
Third, the size of gene sets, even though they represent similar biology, can disproportionately impact significance
scores and complicate interpretation [@karp2021pathway;@mubeen2022influence].
Fourth, tight correlation between each gene's expression within a pathway could exacerbate false positive rates
[@gatti2010heading; @wu2012camera].
Fifth, slight differences in the implementation of ORA tests can impact results in some circumstances
@ziemann2024two.
Lastly, some web-based FEA tools lack longevity.
For example, DAVID 6.8 [@huang2009systematic;@Sherman2022-kg] has been used for over 10,000 publications but
since 2022 has been taken offline, leaving these articles irreproducible.
As web-based tools appear to be the most popular option for FEA [@xie2021popularity;@Wijesooriya2022-we],
we should advocate preservation/archiving of the tool as a Docker image [eg: @Ge2020-lr],
to enable reproducibility into the future @perkel2020challenge.

# Conclusion

So why are problems so pervasive in enrichment analysis?
It's likely a combination of poor researcher training, supervision and peer-review scrutiny.
The design of tools and (low) quality of tool documentation might also play a role.
We also know that inadequate methods have a type of advantage compared to the more rigorous ones
due to researcher preferences to present "significant" findings @smaldino2016natural and reliance upon default settings
even if they are incorrect [@mubeen2022influence;@Wijesooriya2022-we].
Problems 2, 3, 5 and 6 appear to be specific to ORA-based tools, and can be avoided entirely by switching to
FCS tools like GSEA, which has the added benefit of enhanced accuracy in terms of precision and
recall [@kaspi2020mitch;@ziemann2024two].
Although learning and running FCS tools is more difficult and time-consuming, the benefits to the quality of results
are substantial.

The deluge of manuscripts in genomics and other fields places an increasing burden on a limited pool of competent,
voluntary peer-reviewers to a point where editors are struggling to maintain high quality,
consistent peer-review @adam2025peer.
As studies become ever more multi-disciplinary, it is increasingly difficult for editors to
cover all the technical aspects thoroughly, causing mistakes like the above to become widespread.
Ultimately, responsibility for using best practices in enrichment analysis lies with authors.
Avoiding these mistakes will have many benefits, including avoiding wasted resources on unreliable research directions.

# Methods

The example RNA-seq dataset used involves AML3 cells with and without azacytidine treatment
@lund2014dnmt.
This dataset was selected because it represents a typical transcriptomic study; with three
experimental replicates and over 1000 differentially expressed genes.
Read counts were obtained from the DEE2 database using the getDEE2 R package using SRA
project accession SRP038101 as a query @ziemann2019digital.
Genes with average expression above 10 counts per sample were considered detected, and
genes not meeting this criterion were removed from downstream analysis (unless stated).
Differential expression was conducted using DESeq2 v1.48.2 @love2014differential.
Human gene symbols were updated to Ensembl 115 @dyer2025ensembl.
Genes with FDR<0.05 were considered significantly differentially expressed.
Reactome gene sets were downloaded in GMT format on 02/09/2025 @milacic2024reactome, and
these gene sets were used for all subsequent enrichment tests unless otherwise stated.
All ORA tests were conducted using the `fora` function belonging to the `fgsea` package
v1.34.2 @Korotkevich2021-xm.
Minimum gene set size was set to 5 for all subsequent enrichment tests.
Gene sets with FDR<0.05 were considered significantly differentially expressed.

To demonstrate the importance of FDR control (Mistake 1), random sampling of 1000 detected
genes was followed with ORA with and without FDR control.
This was repeated 100 times.
For comparison of the nominal and FDR corrected results, the top 1000 up- and 1000
down-regulated genes by p-value altered by azacitidine treatment underwent separate ORA tests
using a background consisting of the 13,063 detected genes with gene symbols.
The number of significant gene sets (FDR<0.05) from each test were combined for visualisation
using an Euler diagram.

To demonstrate the importance of a suitable background gene list (Mistake 2), random sampling
of 1000 detected genes as the foreground followed by ORA using either the whole genome
background or background consisting of detected genes.
This process was repeated 100 times.
To show the effect of incorrect background on the azacitidine gene profile, 1000 up- and
1000 down-regulated genes selected by p-value underwent ORA using a background including
55,503 genes present in the genome annotation.
These results were compared to those generated using a background consisting of the 13,063
detected genes using an Euler diagram.

To demonstrate the importance of using enrichment scores for interpretation of FEA results,
(Mistake 3 and 4) the `mitch` package v1.20.0 was used for FCS analysis of the differential
expression data @kaspi2020mitch.

To investigate how foreground gene set size impacts the number of ORA results (Mistake 5),
the top N significant up- and down-regulated genes were selected for ORA compared to the
background consisting of detected genes, where N was varied between 50 and 7000.

To demonstrate the effect of combining up and down regulated genes into a single test
(Mistake 6), the top significant 750 genes irrespective of fold change direction were used
as the foreground for an ORA test, and compared with results from the separate approach
with 750 up- and 750 down-regulated genes respectively.

To show the effect of using shallow gene annotations (Mistake 7), various gene set
libraries were extracted from the MSigDB Collection (msigdb.v2025.1.Hs.symbols.gmt)
@liberzon2015molecular.
These gene set libraries were each used for ORA tests with foreground lists of 750 genes as
above.

To show the difference in size of old gene sets (Mistake 8), Reactome gene sets were
extracted from archived MSigDB Collections going back to 2010.
These were compared to the most recent Reactome release (02/09/2025) obtained directly from
the Reactome website.

Instances of poor presentation (Mistake 9) and methods reproducibility (Mistake 10)
were drawn from unpublished notes made during a previous systematic analysis of the
literature @Wijesooriya2022-we.

Analysis was conducted in R 4.5.1 using a Bioconductor Docker image corresponding to
Bioconductor release 3.21.
Analysis code is available from GitHub (https://github.com/markziemann/10mistakes)
and the Docker image is available from DockerHub (https://hub.docker.com/repository/docker/mziemann/10mistakes/general).
These will be archived to Zenodo upon acceptance.

# Bibliography
